---
title: "Distance decay analysis"
author: "Liz Ing-Simmons"
date: "11/12/2018"
output: 
  html_document:
    toc: true
    toc_float: false
    code_folding: hide
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=6, dpi = 300, message = FALSE)
pdf.options(useDingbats = FALSE)
options(stringsAsFactors = FALSE)

# colour_scheme <- c(gd7 = "#648FFF", nc14 = "#785EF0", Tollrm910 ="#DC267F", "3-4h" = "#FE6100", Toll10B = "#FFB000", "mitotic" = "black")
colour_scheme <- c(gd7 = "#648FFF",  Tollrm910 ="#DC267F", Toll10B = "#FFB000", "control" = "black")
# colour_scheme <- c("#648FFF", "#DC267F", "#FFB000")
```

```{r load_packages, cache = FALSE}
library("dplyr")
library("tidyr")
library("ggplot2")
```

The aim of this analysis is to assess distance decay patterns for Hi-C data for gd7, Tollrm9/10, and Toll10B mutant embryos at nc14. 

The distance decay is calculated as the average number of contacts at each distance between bins (using kaic). Here I'm plotting raw distance decay values. 

```{r, fig.width = 7, fig.height = 5, out.width="60%"}
basedir <- "~/cluster/dorsal_ventral/for_paper/"
meta <- read.table(file.path(basedir, "metadata.txt"), sep = "\t", header = TRUE)
meta <- filter(meta, condition != "mitotic")
sample_names <- unique(meta$condition)

distance_decay_files <- file.path(basedir, "data", "hic", "merged", sample_names, "hic",
                                  paste0(sample_names, "_1kb_expected_values_all.txt"))

distance_decays <- lapply(distance_decay_files, readr::read_tsv) %>%
  setNames(sample_names) %>%
  bind_rows(.id = "sample") 
# %>%
#   extract(sample, into = c("sample", "replicate"), regex = "([[:alnum:]]+)_Rep([[:alnum:]]+)") %>%
#   mutate(sample = ifelse(sample == "4h", "3-4h", sample))

distance_decays <- distance_decays %>% 
  dplyr::filter(grepl("-stg10|-nc14|3-4h", sample)) %>% 
  tidyr::separate(sample, sep = "-", into = c("genotype", "stage")) %>% 
  mutate(stage = ifelse(stage == "4h", "3-4h", stage),
         genotype = ifelse(genotype == "3", "control", genotype))

ggplot(distance_decays, aes(x = distance, y = expected, colour = genotype)) +
  geom_line(size = 1) +
  scale_x_log10("Distance", breaks = c(10^4, 10^5, 10^6, 10^7), labels = c("10 kb", "100 kb", "1 Mb", "10 Mb")) +
  scale_y_log10("Expected contacts") +
  scale_color_manual(values = colour_scheme) +
  theme_bw(base_size = 12) +
  facet_wrap(~stage)

```

Here I'm plotting smoothed distance decay, using spline smoothing with `spar = 1` and `tol = 0.0005`, and the derivative of the smoothed curve. This is calculated after taking the log10 of both distances and expected values. This is based on what Benjamin has used, and the values are determined empirically to produce smooth curves that show the overall pattern without noise (and in comparison to published examples). 

```{r, fig.width = 7, fig.height = 3, out.width="60%"}
distance_decays_log10 <- distance_decays %>%
  mutate(distance = log10(distance), expected = log10(expected)) %>%
  dplyr::filter(!(is.infinite(distance) | is.infinite(expected)))
# 
# ggplot(distance_decays_log10, aes(x = distance, y = expected, colour = sample, linetype = replicate)) +
#   geom_line(size = 1) +
#   theme_bw(base_size = 16) +
#   scale_x_continuous("Distance", breaks = c(4, 5, 6, 7), labels = c("10 kb", "100 kb", "1 Mb", "10 Mb")) +
#   scale_y_continuous("Expected contacts (log10)") +
#   scale_color_manual(values = colour_scheme)

predict_deriv <- function(x, y, model){
  model <- smooth.spline(x = x, y = y, spar=1, tol = 0.0005)
  predict(model, x = x, deriv = 1)$y
}

predict_smooth <- function(x, y, model){
  model <- smooth.spline(x = x, y = y, spar=1, tol = 0.0005)
  predict(model, x = x)$y
}

distance_decays_log10_model <- distance_decays_log10 %>%
  group_by(genotype, stage) %>%
  mutate(deriv = predict_deriv(distance, expected),
         smooth = predict_smooth(distance, expected))

ggplot(distance_decays_log10_model, aes(x = distance, y = smooth, colour = genotype)) +
  geom_line(size = 1) +
  theme_bw(base_size = 16) +
  scale_x_continuous("Distance", breaks = c(4, 5, 6, 7), labels = c("10 kb", "100 kb", "1 Mb", "10 Mb")) +
  scale_y_continuous("Expected contacts (smoothed, log10)") +
  scale_color_manual(values = colour_scheme) +
  facet_wrap(~stage)

ggplot(distance_decays_log10_model, aes(x = distance, y = deriv, colour = genotype)) +
  geom_line(size = 1) +
  theme_bw(base_size = 16) +
  scale_x_continuous("Distance", breaks = c(4, 5, 6, 7, 8), labels = c("10 kb", "100 kb", "1 Mb", "10 Mb", "100 Mb")) +
  scale_y_continuous("Derivative of expected contacts") +
  scale_color_manual(values = colour_scheme) +
  facet_wrap(~stage)
```

# nc14

```{r fig.width=6, fig.height=3}
p_smooth <- distance_decays_log10_model %>% 
  dplyr::filter(stage %in% c("nc14", "3-4h")) %>% 
  ggplot(aes(x = distance, y = smooth, colour = genotype, linetype = stage)) +
  geom_line(size = 1) +
  theme_bw(base_size = 12) +
  scale_x_continuous("Distance", breaks = c(3, 4, 5, 6, 7), labels = c("1kb", "10 kb", "100 kb", "1 Mb", "10 Mb")) +
  scale_y_continuous("Expected contacts\n(smoothed, log10)") +
  scale_linetype_manual(values = c("nc14" = "solid", "3-4h" = "dotted")) +
  scale_color_manual(values = colour_scheme) + 
  theme(legend.position = "none", panel.grid.minor = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

p_deriv <- distance_decays_log10_model %>% 
  dplyr::filter(stage %in% c("nc14", "3-4h")) %>% 
  ggplot(aes(x = distance, y = deriv, colour = genotype, linetype = stage)) +
  geom_line(size = 1) +
  theme_bw(base_size = 12) +
  scale_x_continuous("Distance", breaks = c(3, 4, 5, 6, 7), labels = c("1kb", "10 kb", "100 kb", "1 Mb", "10 Mb")) +
  scale_y_continuous("Derivative of\nexpected contacts") +
  scale_color_manual(values = colour_scheme) + 
  scale_linetype_manual(values = c("nc14" = "solid", "3-4h" = "dotted")) +
  theme(legend.position = "none", panel.grid.minor = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

legend_b <- cowplot::get_legend(
  p_smooth + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom")
)


p_nc14 <- cowplot::plot_grid(cowplot::plot_grid(p_smooth, p_deriv), 
                        legend_b, ncol = 1, rel_heights = c(1, .2))
p_nc14

pdf(file = "../figures/figure_3_panels/nc14_distance_decay.pdf",
    width = 6, height = 3)
p_nc14
dev.off()
```

# Stage 10

```{r fig.width=5, fig.height=2.5}
p_smooth_stg10 <- distance_decays_log10_model %>% 
  dplyr::filter(stage == "stg10") %>% 
  ggplot(aes(x = distance, y = smooth, colour = genotype)) +
  geom_line(size = 1) +
  theme_bw(base_size = 10) +
  scale_x_continuous("Distance", breaks = c(3, 4, 5, 6, 7), labels = c("1kb", "10 kb", "100 kb", "1 Mb", "10 Mb")) +
  scale_y_continuous("Expected contacts\n(smoothed, log10)") +
  scale_color_manual(values = colour_scheme) + 
  theme(legend.position = "none", panel.grid.minor = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

p_deriv_stg10 <- distance_decays_log10_model %>% 
  dplyr::filter(stage == "stg10") %>% 
  ggplot(aes(x = distance, y = deriv, colour = genotype)) +
  geom_line(size = 1) +
  theme_bw(base_size = 10) +
  scale_x_continuous("Distance", breaks = c(3, 4, 5, 6, 7), labels = c("1kb", "10 kb", "100 kb", "1 Mb", "10 Mb")) +
  scale_y_continuous("Derivative of\nexpected contacts") +
  scale_color_manual(values = colour_scheme) + 
  theme(legend.position = "none", panel.grid.minor = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

legend_b <- cowplot::get_legend(
  p_smooth_stg10 + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom")
)


p_stg10 <- cowplot::plot_grid(cowplot::plot_grid(p_smooth, p_deriv), 
                        legend_b, ncol = 1, rel_heights = c(1, .2))
p_stg10

pdf(file = "../figures/figure_3_panels/stg10_distance_decay.pdf",
    width = 6, height = 3)
p_stg10
dev.off()
```


# Micro-C

```{r}
sample_names <- c("control", "gd7")

distance_decay_files <- file.path(basedir, "data", "micro-c", "merged", sample_names, "hic",
                                  paste0(sample_names, "_100bp_expected_values_all.txt"))

distance_decays <- lapply(distance_decay_files, readr::read_tsv) %>%
  setNames(sample_names) %>%
  bind_rows(.id = "sample") 

ggplot(distance_decays, aes(x = distance, y = expected, colour = sample)) +
  geom_line(size = 1) +
  scale_x_log10("Distance", breaks = c(10^2, 10^3, 10^4, 10^5, 10^6, 10^7), 
                labels = c("100 bp", "1 kb", "10 kb", "100 kb", "1 Mb", "10 Mb")) +
  scale_y_log10("Expected contacts") +
  scale_color_manual(values = colour_scheme) +
  theme_bw(base_size = 12) 

distance_decays_log10 <- distance_decays %>%
  mutate(distance = log10(distance), expected = log10(expected)) %>%
  dplyr::filter(!(is.infinite(distance) | is.infinite(expected)))

distance_decays_log10_model <- distance_decays_log10 %>%
  group_by(sample) %>%
  mutate(deriv = predict_deriv(distance, expected),
         smooth = predict_smooth(distance, expected))

ggplot(distance_decays_log10_model, aes(x = distance, y = smooth, colour = sample)) +
  geom_line(size = 1) +
  theme_bw(base_size = 16) +
  scale_x_continuous("Distance", breaks = 2:8, 
                labels = c("100 bp", "1 kb", "10 kb", "100 kb", "1 Mb", "10 Mb", "100 Mb")) +
  scale_y_continuous("Expected contacts (smoothed, log10)") +
  scale_color_manual(values = colour_scheme)

ggplot(distance_decays_log10_model, aes(x = distance, y = deriv, colour = sample)) +
  geom_line(size = 1) +
  theme_bw(base_size = 16) +
  scale_x_continuous("Distance", breaks = 2:8, 
                labels = c("100 bp", "1 kb", "10 kb", "100 kb", "1 Mb", "10 Mb", "100 Mb")) +
  scale_y_continuous("Derivative of expected contacts") +
  scale_color_manual(values = colour_scheme)
```
```{r fig.width=6, fig.height=3}
p_smooth <- distance_decays_log10_model %>% 
  ggplot(aes(x = distance, y = smooth, colour = sample)) +
  geom_line(size = 1) +
  theme_bw(base_size = 10) +
  scale_x_continuous("Distance", breaks = 2:8, 
                labels = c("100 bp", "1 kb", "10 kb", "100 kb", "1 Mb", "10 Mb", "100 Mb")) +
  scale_y_continuous("Expected contacts\n(smoothed, log10)") +
  scale_color_manual(values = colour_scheme) + 
  theme(legend.position = "none", panel.grid.minor = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

p_deriv <- distance_decays_log10_model %>% 
  ggplot(aes(x = distance, y = deriv, colour = sample)) +
  geom_line(size = 1) +
  theme_bw(base_size = 10) +
  scale_x_continuous("Distance", breaks = 2:8, 
                labels = c("100 bp", "1 kb", "10 kb", "100 kb", "1 Mb", "10 Mb", "100 Mb")) +
  scale_y_continuous("Derivative of\nexpected contacts") +
  scale_color_manual(values = colour_scheme) + 
  theme(legend.position = "none", panel.grid.minor = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

legend_b <- cowplot::get_legend(
  p_smooth + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom")
)


p_microc <- cowplot::plot_grid(cowplot::plot_grid(p_smooth, p_deriv), 
                        legend_b, ncol = 1, rel_heights = c(1, .2))
p_microc

pdf(file = "../figures/figure_5_panels/microc_distance_decay.pdf",
    width = 6, height = 3)
p_microc
dev.off()
```



```{r fig.width=6, fig.height=4}
legend_b <- cowplot::get_legend(
  p_smooth_stg10 + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom")
)


cowplot::plot_grid(cowplot::plot_grid(p_smooth, p_deriv, p_smooth_stg10, p_deriv_stg10, 
                                      nrow = 2, align = "vh", axis = "lr"),
                   legend_b, nrow = 2, rel_heights = c(2, .2))

pdf(file = "../figures/figure_5_panels/microc_stg10_distance_decay.pdf",
    width = 6, height = 4)
cowplot::plot_grid(cowplot::plot_grid(p_smooth, p_deriv, p_smooth_stg10, p_deriv_stg10, 
                                      nrow = 2, align = "vh", axis = "lr"),
                   legend_b, nrow = 2, rel_heights = c(2, .2))
dev.off()
```


# Session info

This report was generated at `r format(Sys.time(), "%X, %a %b %d %Y")`. 

```{r}
devtools::session_info()
```
