---
title: "Resolution analysis"
author: "Liz Ing-Simmons"
date: "06/09/2019"
output: 
  html_document:
    code_folding: hide
---

```{r global_options, echo=FALSE}
short=TRUE #if short==TRUE, do not echo code chunks
debug=FALSE
knitr::opts_chunk$set(fig.width=6, fig.height=6, dpi = 300,
             echo=!short, warning=debug, message=debug, error= FALSE)
pdf.options(useDingbats = FALSE)
options(stringsAsFactors = FALSE)
```

```{r load_packages, cache = FALSE}
library("dplyr")
library("tidyr")
library("readr")
library("ggplot2")

read_bed <- function(fn){
  read_tsv(fn, col_names = c("chr", "start", "end", "name", "score", "strand"))
}
```

```{r load_bed_files, cache=TRUE}
## Load .bed files

basedir <- here::here()
sample_names <- c("mitotic", "nc14", "3-4h", 
                  "control-nc14", "gd7-nc14", "Toll10B-nc14", "Tollrm910-nc14",
                  "control-stg10", "gd7-stg10", "Toll10B-stg10", "Tollrm910-stg10")
resolutions <- c("1kb", "2kb", "5kb", "10kb", "25kb")
# , "50kb", "100kb", "250kb", 
#                  "500kb","1mb")

sample_names_expanded <- rep(sample_names, each = length(resolutions))
resolutions_expanded <- rep(resolutions, length(sample_names))

files <- file.path(basedir, "data", "hic", "merged", sample_names_expanded, "hic",
                   paste0(sample_names_expanded, "_", resolutions_expanded, 
                          "_marginals.bed"))
files <- files[file.exists(files)]
binned_beds <- lapply(files, read_bed)
names(binned_beds) <- basename(files)
binned_beds_df <- bind_rows(binned_beds, .id = "file") %>%
  separate(file, into = c("sample", "resolution", "rest"), sep = "_", extra = "merge") %>%
  select(-name, -strand, -rest)
```

## Lieberman-Aiden lab resolution assessment

Rao, Huntley et al 2014 assess the number of bins with > 1000 contacts at different resolutions, and take the "maximum" resolution of the sample to be the resolution at which at least 80% of bins have > 1000 contacts. 

```{r}
binned_beds_df %>%
  mutate(resolution = factor(resolution, levels = resolutions, ordered = TRUE)) %>%
  group_by(sample, resolution) %>%
  summarise(percent = 100* sum(score > 1000) / n()) %>%
  spread(resolution, percent) %>%
  knitr::kable(digits = 1)
```

Clemens' samples reach the 80% threshold at 2 kb, mine reach this threshold at 5 kb. 


## Session info

This report was generated at `r format(Sys.time(), "%X, %a %b %d %Y")`. 

```{r}
sessionInfo()
```
